---
title: "LTR-TE Dynamics — GLMM Analysis with TE Family Effects"
author: "Marie Kratka"
date: "`r Sys.Date()`"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = TRUE,
  message = TRUE,
  fig.width  = 8,
  fig.height = 5
)
```

# Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(glmmTMB)
library(broom.mixed)
library(DHARMa)
library(knitr)
library(here)
library(patchwork)
library(svglite)
library(emmeans)
```

------------------------------------------------------------------------

# Load and prepare data

The mega-table has one row per LTR-TE element (intact, partial, or
solo_LTR), with element-level, window-level, chromosome-level, and
species-level metadata.

```{r load-data}
df_raw <- read_tsv(here("data", "df_for_linear_models.tsv"), show_col_types = FALSE)
```

```{r prepare-factors}
df_raw <- df_raw %>%
  mutate(
    # monocentric as reference level
    arch = factor(Centromere_architecture,
                  levels = c("monocentric", "holocentric")),
    Family       = factor(Family),
    Ordo         = factor(Ordo),
    Clade        = factor(Clade),
    Genus        = factor(Genus),
    Species      = factor(Species),
    Class        = factor(Class),
    element_type = factor(element_type, levels = c("intact", "partial", "solo_LTR")),
    # Log chromosome size
    log_chr_length = log10(chr_length / 1e6)
  )
```

```{r extract-te-family}
# TE family name (e.g. "Angela", "Athila", "CRM", "Tekay") from DANTE Class string
# Class column typically looks like: "Ty1/copia|Angela" or "Ty3/gypsy|chromovirus|Tekay|CRM"
# Adjust parsing to match your actual Class format

df_raw <- df_raw %>%
  mutate(
    # Extract the last element after the last "|" — this is typically the family name
    TE_family = str_extract(Class, "[^|]+$"),
    TE_family = factor(TE_family),
    # Superfamily: Ty1-copia vs Ty3-gypsy
    superfamily = case_when(
      grepl("Ty1/copia", Class, ignore.case = TRUE) ~ "Ty1-copia",
      grepl("Ty3/gypsy", Class, ignore.case = TRUE) ~ "Ty3-gypsy",
      TRUE ~ "other"
    ),
    superfamily = factor(superfamily)
  )

# Verify parsing
cat("TE family levels:\n")
print(sort(table(df_raw$TE_family), decreasing = TRUE))

cat("\nSuperfamily levels:\n")
print(table(df_raw$superfamily))
```

## Sensitivity filter

```{r sensitivity-filter}
sensitivity_exclude <- c(
  "Hordeum_vulgare",
  "Secale_cereale",
  "Triticum_aestivum",
  "Triticum_monococcum",
  "Rhynchospora_pubera"
)

df_full      <- df_raw
df_sensitive <- df_raw %>% filter(!Species %in% sensitivity_exclude)

cat("Full dataset:       ", n_distinct(df_full$Species),      "species\n")
cat("Sensitivity subset: ", n_distinct(df_sensitive$Species), "species\n")
```

# load family palette
```{r family_palette}
family_palette <- c(
  "Convolvulaceae" = "#f3b674",
  "Poaceae"        = "#a6f3a6",
  "Juncaceae"      = "#ff41c8",
  "Cyperaceae"     = "#916f6f",
  "Melanthiaceae"  = "#37abc8"
)
```

------------------------------------------------------------------------

# Aggregate to chromosome level

Used for hypotheses tested at chromosome level (H1–H4).

```{r aggregate-chromosome}
aggregate_chrom <- function(df) {
  df %>%
    group_by(Species, chr, chr_length, arch,
             Family, Ordo, Clade, Genus) %>%
    summarise(
      n_intact  = sum(element_type == "intact"),
      n_partial = sum(element_type == "partial"),
      n_solo    = sum(element_type == "solo_LTR"),
      n_total   = n(),

      solo_ratio = n_solo / (n_intact + n_solo),

      mean_LTR_identity = mean(LTR_identity[element_type == "intact"], na.rm = TRUE) / 100,

      TE_density = sum(end - start) / first(chr_length),
      crm_prop = sum(grepl("CRM", Class, ignore.case = TRUE)) / n_total,
      gypsy_prop = sum(grepl("Ty3-gypsy", superfamily, ignore.case = TRUE)) / n_total,

      .groups = "drop"
    ) %>%
    mutate(
      log_chr_length = log10(chr_length / 1e6),
      log_chr_length_c = log_chr_length - mean(log_chr_length, na.rm = TRUE),

      # Smithson & Verkuilen (2006) boundary correction for beta regression
      solo_ratio_adj   = (solo_ratio        * (n_total - 1) + 0.5) / n_total,
      ltr_identity_adj = (mean_LTR_identity * (n_total - 1) + 0.5) / n_total,
      TE_density_adj   = (TE_density        * (n_total - 1) + 0.5) / n_total,
      crm_prop_adj     = (crm_prop          * (n_total - 1) + 0.5) / n_total,
      gypsy_prop_adj   = (gypsy_prop        * (n_total -1) + 0.5) / n_total
    ) %>%
    filter(
      !is.na(solo_ratio),
      !is.na(mean_LTR_identity),
      !is.na(TE_density),
      n_intact >= 5
    )
}

chrom_full      <- aggregate_chrom(df_full)
chrom_sensitive <- aggregate_chrom(df_sensitive)

cat("Chromosomes (full):        ", nrow(chrom_full),      "\n")
cat("Chromosomes (sensitivity): ", nrow(chrom_sensitive), "\n")
```

```{r inspect-chrom}
chrom_full %>%
  group_by(arch) %>%
  summarise(
    n_species         = n_distinct(Species),
    n_chr             = n(),
    mean_solo_ratio   = mean(solo_ratio,        na.rm = TRUE),
    mean_ltr_identity = mean(mean_LTR_identity, na.rm = TRUE),
    mean_TE_density   = mean(TE_density,        na.rm = TRUE),
    .groups = "drop"
  ) %>%
  kable(digits = 4, caption = "Chromosome-level summary by architecture")
```

------------------------------------------------------------------------

# Aggregate to window level

Used for hypotheses tested at the window level (H1, H3).
Keep distinct LTR-TE families separate to allow family-level effects in the models.


```{r aggregate-window}

aggregate_window <- function(df) {
  
  window_totals <- df %>%
    group_by(Species, chr, bin_ID, chr_length, arch, 
             Family, Ordo, Clade, Genus) %>%
    summarise(
      all_intact_count_in_window = sum(element_type == "intact"),
      all_disrupted_count_in_window = sum(element_type == "partial"),
      all_solo_count_in_window = sum(element_type == "solo_LTR"),
      all_total = n(),
      all_mean_LTR_identity_in_window = mean(LTR_identity[element_type == "intact"], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      all_solo_intact_ratio_in_window = ifelse(all_intact_count_in_window + all_solo_count_in_window > 0,
                        all_solo_count_in_window / (all_intact_count_in_window + all_solo_count_in_window),
                        NA_real_),
      all_TE_density_in_window = all_intact_count_in_window / (0.05*chr_length),  # 5% window size
      log_chr_length = log10(chr_length / 1e6),
      log_chr_length_c = log_chr_length - mean(log_chr_length, na.rm = TRUE),
      
      # Verkuilen adjustment
      all_solo_intact_ratio_adj = ifelse(!is.na(all_solo_intact_ratio_in_window),
                        (all_solo_intact_ratio_in_window * (all_total - 1) + 0.5) / all_total,
                        NA_real_)
  )
  
}

window_sensitive <- aggregate_window(df_sensitive)
window_full <- aggregate_window(df_full)
cat("Windows included sensitive:", nrow(window_sensitive), "\n")
cat("Windows included full:", nrow(window_full), "\n")

```

------------------------------------------------------------------------

# Filter to element level

TE_density_adj   = (TE_density        * (n_total - 1) + 0.5) / n_total,
```{r filter-element}

prep_element_data <- function(df, min_family_n = 100) {
  out <- df %>%
    filter(
      element_type == "intact",
      !is.na(LTR_identity),
      !is.na(log_chr_length),
      !is.na(arch),
      !is.na(Class)
    ) %>%
    mutate(
      LTRi_prop = (LTR_identity/100 * (LTR_length - 1) + 0.5) / LTR_length,
      log_chr_length_c = log_chr_length - mean(log_chr_length, na.rm = TRUE),
      log_LTR_length = log10(LTR_length + 1e-6)
    )

  cat("Intact elements:", nrow(out), "\n")
  cat("TE families present:\n")
  print(sort(table(out$Class), decreasing = TRUE))

  family_counts <- out %>% count(Class) %>% filter(n >= min_family_n)
  cat("\nFamilies with >=", min_family_n, "elements:", nrow(family_counts), "\n")

  out <- out %>%
    filter(Class %in% family_counts$Class) %>%
    mutate(Class = droplevels(Class))

  cat("Elements after family filter:", nrow(out), "\n")
  out
}

element_full      <- prep_element_data(df_full)
element_sensitive <- prep_element_data(df_sensitive)

```





------------------------------------------------------------------------
# H1a: overall TE density

```{r h1a_additive}
h1a_additive <- glmmTMB(TE_density_adj ~ arch + log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = chrom_sensitive)
summary(h1a_additive)
```

```{r h1a_interaction}
h1a_interaction <- glmmTMB(TE_density_adj ~ arch * log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = chrom_sensitive)
summary(h1a_interaction)
```
Additive model has better BIC.
Architecture has no effect, chromosome size associated with increased TE abundance.

Is there a difference on chrom_full?
```{r h1a_full}
h1a_full <- glmmTMB(TE_density_adj ~ arch + log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = chrom_full)
summary(h1a_full)
```
Nope.

Now visualize the distribution

```{r te_abundance_visualization}
p_density <- chrom_sensitive %>%
  ggplot(aes(x = log_chr_length, y = TE_density)) +
  stat_density_2d(
    aes(fill = arch, alpha = after_stat(level)),
    geom = "polygon"
  ) +
  scale_fill_manual(values = c("monocentric" = "gray20",
                               "holocentric" = "blue")) +
  scale_alpha_continuous(range = c(0.05, 0.4)) +
  labs(x = "log10(chromosome length)", y = "LTR-TE density",
       fill = "Architecture") +
  guides(alpha = "none") +
  theme_bw() +
  theme(text = element_text(family = "Arial"))

p_points <- chrom_sensitive %>%
  ggplot(aes(x = log_chr_length, y = TE_density)) +
  geom_point(aes(shape = arch),
             color = "black", size = 1.5) +
  geom_point(aes(color = Family, shape = arch),
             size = 1.2, alpha = 0.8) +
  scale_color_manual(values = family_palette) +
  scale_shape_manual(values = c("holocentric" = 17,
                                "monocentric" = 16)) +
  labs(x = "log10(chromosome length)", y = "LTR-TE density",
       color = "Family", shape = "Architecture") +
  theme_bw() +
  theme(text = element_text(family = "Arial"))

p_combined <- p_density + p_points +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
        text = element_text(family = "Arial"))
ggsave("../../data/figures/GLMM/te_density_vs_chr_size.svg", plot = p_combined,
       width = 8, height = 5, device = svglite)
p_combined

```

And export the effects estimation table

```{r h1a_export}
tidy(h1a_additive, effects = "fixed", conf.int = TRUE) |>
write_csv("../../data/tables/glmm/h1a.csv")
```

------------------------------------------------------------------------

# H1b: CRM proportions

not enough elements in individual windows, calculate on chromosome level

```{r h1b_additive}
h1b_additive <- glmmTMB(crm_prop_adj ~ arch + log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = chrom_sensitive)
summary(h1b_additive)
```

```{r h1b_interaction}
h1b_interaction <- glmmTMB(crm_prop_adj ~ arch * log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = chrom_sensitive)
summary(h1b_interaction)
```
Interaction is better.
In monocentrics, CRM goes down with chr size.
There is fewer CRM in holocentrics.

```{r h1b_full}
h1b_full <- glmmTMB(crm_prop_adj ~ arch * log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = chrom_full)
summary(h1b_full)
```
Same result for full dataset

```{r CRM_abundance_visualization}
p_density <- chrom_sensitive %>%
  ggplot(aes(x = log_chr_length, y = crm_prop)) +
  stat_density_2d(
    aes(fill = arch, alpha = after_stat(level)),
    geom = "polygon"
  ) +
  scale_fill_manual(values = c("monocentric" = "gray20",
                               "holocentric" = "blue")) +
  scale_alpha_continuous(range = c(0.05, 0.4)) +
  labs(x = "log10(chromosome length)", y = "CRM proportion",
       fill = "Architecture") +
  guides(alpha = "none") +
  theme_bw() +
  theme(text = element_text(family = "Arial"))

p_points <- chrom_sensitive %>%
  ggplot(aes(x = log_chr_length, y = crm_prop)) +
  geom_point(aes(shape = arch),
             color = "black", size = 1.5) +
  geom_point(aes(color = Family, shape = arch),
             size = 1.2, alpha = 0.8) +
  scale_color_manual(values = family_palette) +
  scale_shape_manual(values = c("holocentric" = 17,
                                "monocentric" = 16)) +
  labs(x = "log10(chromosome length)", y = "CRM proportion",
       color = "Family", shape = "Architecture") +
  theme_bw() +
  theme(text = element_text(family = "Arial"))

p_combined <- p_density + p_points +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
        text = element_text(family = "Arial"))
ggsave("../../data/figures/GLMM/crmprop_vs_chr_size.svg", plot = p_combined,
       width = 8, height = 5, device = svglite)
p_combined

```

Export results
```{r h1b_export}
tidy(h1b_interaction, effects = "fixed", conf.int = TRUE) |>
write_csv("../../data/tables/glmm/h1b.csv")
```


# H1c Gypsy proportion

```{r h1c_additive}
h1c_additive <- glmmTMB(gypsy_prop_adj ~ arch + log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = chrom_sensitive)
summary(h1c_additive)
```

```{r h1c_interaction}
h1c_interaction <- glmmTMB(gypsy_prop_adj ~ arch * log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = chrom_sensitive)
summary(h1c_interaction)
```
Additive is better.
The gypsy proportion goes up in larger chromosomes (both groups).
No difference between mono and holocentrics.

```{r h1c_full}
h1c_full <- glmmTMB(gypsy_prop_adj ~ arch + log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = chrom_full)
summary(h1c_full)
```
```{r gypsy_abundance_visualization}
p_density <- chrom_sensitive %>%
  ggplot(aes(x = log_chr_length, y = gypsy_prop)) +
  stat_density_2d(
    aes(fill = arch, alpha = after_stat(level)),
    geom = "polygon"
  ) +
  scale_fill_manual(values = c("monocentric" = "gray20",
                               "holocentric" = "blue")) +
  scale_alpha_continuous(range = c(0.05, 0.4)) +
  labs(x = "log10(chromosome length)", y = "Ty3-gypsy proportion",
       fill = "Architecture") +
  guides(alpha = "none") +
  theme_bw() +
  theme(text = element_text(family = "Arial"))

p_points <- chrom_sensitive %>%
  ggplot(aes(x = log_chr_length, y = gypsy_prop)) +
  geom_point(aes(shape = arch),
             color = "black", size = 1.5) +
  geom_point(aes(color = Family, shape = arch),
             size = 1.2, alpha = 0.8) +
  scale_color_manual(values = family_palette) +
  scale_shape_manual(values = c("holocentric" = 17,
                                "monocentric" = 16)) +
  labs(x = "log10(chromosome length)", y = "Ty-3 gypsy proportion",
       color = "Family", shape = "Architecture") +
  theme_bw() +
  theme(text = element_text(family = "Arial"))

p_combined <- p_density + p_points +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
        text = element_text(family = "Arial"))
ggsave("../../data/figures/GLMM/gypsyprop_vs_chr_size.svg", plot = p_combined,
       width = 8, height = 5, device = svglite)
p_combined

```


------------------------------------------------------------------------
# H2a: LTR identity

without family

```{r h2a_additive}
h2a_additive <- glmmTMB(LTRi_prop ~ arch + log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = element_sensitive)
summary(h2a_additive)
```


```{r h2a_interaction}
h2a_interaction <- glmmTMB(LTRi_prop ~ arch*log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = element_sensitive)
summary(h2a_interaction)
```
Interaction is better
with family

# H2b family as random 

```{r h2b_interaction}

h2b_interaction <- glmmTMB(LTRi_prop ~ arch * log_chr_length_c +
          (1 | Ordo/Family/Genus/Species) + (1 | Class),
        family = beta_family(),
        dispformula = ~arch,
        data = element_sensitive)

summary(h2b_interaction)
```
with full dataset

```{r h2b_full}


h2b_full <- glmmTMB(LTRi_prop ~ arch * log_chr_length_c +
          (1 | Ordo/Family/Genus/Species) + (1 | Class),
        family = beta_family(),
        dispformula = ~arch,
        data = element_full)

summary(h2b_full)
```








```{r h2b_visualization}
p_density <- chrom_sensitive %>%
  ggplot(aes(x = log_chr_length, y = mean_LTR_identity)) +
  stat_density_2d(
    aes(fill = arch, alpha = after_stat(level)),
    geom = "polygon"
  ) +
  scale_fill_manual(values = c("monocentric" = "gray20",
                               "holocentric" = "blue")) +
  scale_alpha_continuous(range = c(0.05, 0.4)) +
  labs(x = "log10(chromosome length)", y = "LTR identity (%)",
       fill = "Architecture") +
  guides(alpha = "none") +
  theme_bw() +
  theme(text = element_text(family = "Arial"))

p_points <- chrom_sensitive %>%
  ggplot(aes(x = log_chr_length, y = mean_LTR_identity)) +
  geom_point(aes(shape = arch),
             color = "black", size = 1.5) +
  geom_point(aes(color = Family, shape = arch),
             size = 1.2, alpha = 0.8) +
  scale_color_manual(values = family_palette) +
  scale_shape_manual(values = c("holocentric" = 17,
                                "monocentric" = 16)) +
  labs(x = "log10(chromosome length)", y = "LTR identity (%)",
       color = "Family", shape = "Architecture") +
  theme_bw() +
  theme(text = element_text(family = "Arial"))

p_combined <- p_density + p_points +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
        text = element_text(family = "Arial"))
ggsave("../../data/figures/GLMM/ltr_identity_vs_chr_size.svg", plot = p_combined,
       width = 8, height = 5, device = svglite)
p_combined

```

# H2c families differ in their LTR identity (interaction)
effect of the family

```{r}

h2c_interaction <- glmmTMB(LTRi_prop ~ arch * log_chr_length_c + arch*TE_family +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = element_sensitive)
summary(h2c_interaction)
```

```{r}
emm <- emmeans(h2c_interaction, ~ arch | TE_family, type = "response")

emm_df <- as.data.frame(emm)

# Calculate holo - mono difference per family
family_diffs <- emm_df %>%
  select(TE_family, arch, response) %>%
  pivot_wider(names_from = arch, values_from = response) %>%
  mutate(diff = holocentric - monocentric)

# Join back and use diff for ordering
emm_df <- emm_df %>%
  left_join(family_diffs %>% select(TE_family, diff), by = "TE_family")

ggplot(emm_df, aes(x = response, y = reorder(TE_family, diff),
                    color = arch, shape = arch)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = asymp.LCL, xmax = asymp.UCL), height = 0.2) +
  scale_color_manual(values = c("monocentric" = "black", "holocentric" = "#0000fff5")) +
  labs(x = "Estimated LTR identity (back-transformed)",
       y = "TE Family",
       color = "Architecture", shape = "Architecture",
       title = "Estimated LTR identity by TE family and architecture",
       caption = "EMMs at mean chromosome size; error bars = 95% CI") +
  theme_bw() +
  theme(legend.position = "bottom",
        text = element_text(family = "Arial"))

ggsave("../../data/figures/GLMM/ltr_identity_by_family_arch.svg", width = 4, height = 4, device = svglite)
```


-------------------------------------------------------------

# H2d - the effect of LTR length

```{r h2d_additive}
h2d_additive <- glmmTMB(LTRi_prop ~ arch * log_chr_length_c + log_LTR_length +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = element_full)
summary(h2d_additive)

```
```{r h2d_with_family}
h2d_additive_family <- glmmTMB(LTRi_prop ~ arch * log_chr_length_c + TE_family + arch:TE_family +
               log_LTR_length + arch:log_LTR_length +
               (1 | Ordo/Family/Genus/Species),
               family = beta_family(),
               dispformula = ~arch,
               data = element_sensitive)
summary(h2d_additive_family)
```
```{r length_plus_fam_emm}
emm <- emmeans(h2d_additive_family, ~ arch | TE_family, type = "response")

emm_df <- as.data.frame(emm)

# Calculate holo - mono difference per family
family_diffs <- emm_df %>%
  select(TE_family, arch, response) %>%
  pivot_wider(names_from = arch, values_from = response) %>%
  mutate(diff = holocentric - monocentric)

# Join back and use diff for ordering
emm_df <- emm_df %>%
  left_join(family_diffs %>% select(TE_family, diff), by = "TE_family")

ggplot(emm_df, aes(x = response, y = reorder(TE_family, diff),
                    color = arch, shape = arch)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = asymp.LCL, xmax = asymp.UCL), height = 0.2) +
  scale_color_manual(values = c("monocentric" = "black", "holocentric" = "#0000fff5")) +
  labs(x = "Estimated LTR identity (back-transformed)",
       y = "TE Family",
       color = "Architecture", shape = "Architecture",
       title = "Estimated LTR identity by TE family and architecture",
       caption = "EMMs at mean chromosome size; error bars = 95% CI") +
  theme_bw() +
  theme(legend.position = "bottom",
        text = element_text(family = "Arial"))
ggsave("../../data/figures/GLMM/lengths_corrected_ltr_identity_by_family_arch.svg", width = 4, height = 4, device = svglite)
```


```{r}
h2d_interaction <- glmmTMB(LTRi_prop ~ arch * log_chr_length_c + arch*log_LTR_length +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = element_sensitive)
summary(h2d_interaction)

```


```{r}
element_full %>%
  slice_sample(n=10000) %>%
  ggplot(aes(x = log_LTR_length, y = LTRi_prop, color = arch)) +
  geom_point(alpha = 0.1, size = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ TE_family) +
  scale_color_manual(values = c("monocentric" = "gray30", "holocentric" = "blue")) +
  labs(x = "log10(LTR length)", y = "LTR identity",
       color = "Architecture") +
  theme_bw() +
  theme(text = element_text(family = "Arial"),
        legend.position = "bottom")
```

```{r}
element_full %>%
  mutate(LTR_length_bin = cut(LTR_length, breaks = c(0, 100, 200, 500, 1000, 2000, Inf))) %>%
  ggplot(aes(x = LTR_identity, fill = arch)) +
  geom_histogram(position = "dodge", bins = 50) +
  xlim(c(98,100)) +
  facet_wrap(~ LTR_length_bin) +
  theme_bw()
```
```{r}
element_full %>%
  ggplot(aes(x = arch, y = LTR_length, fill = arch)) +
  geom_violin() +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_fill_manual(values = c("monocentric" = "gray30", "holocentric" = "blue")) +
  scale_y_log10() +
  facet_wrap(~ TE_family) +
  labs(x = "Architecture", y = "LTR length (bp, log scale)") +
  theme_bw() +
  theme(legend.position = "none",
        text = element_text(family = "Arial"))
ggsave("../../data/figures/GLMM/ltr_length_in_families.svg", width = 8, height = 8, device = svglite)

```

```{r}
ltr_length_model_fam <- glmmTMB(log_LTR_length ~ arch * TE_family + 
                                  (1 | Family/Genus/Species),
                                family = gaussian(),
                                data = element_full)
summary(ltr_length_model_fam)
```


```{r}
element_full %>%
  group_by(TE_family) %>%
  group_map(~ {
    m <- glmmTMB(log_LTR_length ~ arch + (1 | Family/Genus/Species),
                 family = gaussian(), data = .x)
    broom.mixed::tidy(m, effects = "fixed") %>%
      filter(term == "archholocentric") %>%
      mutate(family = .y$TE_family)
  }) %>%
  bind_rows()
```

-------------------------------------------------------------

# H3a - overall solo ratio


```{r h3a_additive}

h3a_additive <- glmmTMB(all_solo_intact_ratio_adj ~ arch + log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = window_sensitive)

summary(h3a_additive)
```
```{r h3a_interaction}
h3a_interaction <- glmmTMB(all_solo_intact_ratio_adj ~ arch * log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = window_sensitive)
summary(h3a_interaction)
```



# H3b - solo ratio on chromosomal level
```{r h3b_additive}
h3b_additive <- glmmTMB(solo_ratio_adj ~ arch + log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = chrom_sensitive)
summary(h3b_additive)
```

```{r h3b_interaction}
h3b_interaction <- glmmTMB(solo_ratio_adj ~ arch * log_chr_length_c +
          (1 | Ordo/Family/Genus/Species),
        family = beta_family(),
        dispformula = ~arch,
        data = chrom_sensitive)
summary(h3b_interaction)
```
```{r save_h3b}

tidy(h3b_interaction, effects = "fixed", conf.int = TRUE) |>
  write_csv("../../data/tables/glmm/h3b_interaction.csv")
```

# solo ratio slope
```{r solo_slope}
slope <- chrom_sensitive %>%
  group_by(Family, arch) %>%
  filter(n() >= 10) %>%  # only groups with enough data
  do(tidy(lm(solo_ratio ~ log_chr_length, data = .))) %>%
  filter(term == "log_chr_length") %>%
  select(Family, arch, estimate, std.error, p.value) %>%
  arrange(arch, estimate)

write_csv(slope, "../../data/tables/glmm/solo_family_slope.csv")
slope

```


Export linear model results
```{r export}

model_summary <- function(model) {
  tidy(model, effects = "fixed") %>%
    mutate(
      call = deparse1(model$call)
    )
}

model_random <- function(model) {
  tidy(model, effects = "ran_pars") %>%
    mutate(
      call = deparse1(model$call)
    )
}

model_list <- list(
  h1a_additive      = h1a_additive,
  h1c_additive      = h1c_additive,
  h1b_interaction   = h1b_interaction,
  h2a_interaction   = h2a_interaction,
  h2b_interaction   = h2b_interaction,
  h2c_interaction   = h2c_interaction,
  h2d_interaction   = h2d_interaction,
  h3a_interaction   = h3a_interaction,
  h3b_additive      = h3b_additive,
  h3b_interaction   = h3b_interaction
)

results <- map_dfr(model_list, model_summary, .id = "model_name")
random  <- map_dfr(model_list, model_random,  .id = "model_name")
fit     <- map_dfr(model_list, glance,         .id = "model_name")

write_csv(results, "../../data/tables/glmm/glmm_fixed_effects.csv")
write_csv(random,  "../../data/tables/glmm/glmm_random_effects.csv")
write_csv(fit,     "../../data/tables/glmm/glmm_fit_stats.csv")

```

